---
title: "Mapping Cyclist Safety in New York City"
description: |
  Is there a relationship between bike path access and cyclist injuries? 
preview: preview.jpg
base_url: https://jasonbixonblog.netlify.com
author:
  - name: Jason Bixon
    url: https://jasonbixon.netlify.com
    affiliation: Merkle Inc.
    affiliation_url: https://merkleinc.com
date: 09-11-2019
repository_url: https://github.com/jbixon13/Radix-blog
output:
  radix::radix_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE, echo=FALSE, warning=FALSE}
library(tidyverse)
library(forcats)
library(lubridate)
library(plotly)
library(scales)
library(forecast)
library(leaflet)
library(leaflet.extras)
library(mapdeck)
library(sf)
library(htmltools)
library(nycgeo)
library(gganimate)
library(gifski)
library(sjPlot)
```

```{r data_read, message=FALSE, echo=FALSE}
# read in datasets 
NYPD <- read_csv('data/nypd-motor-vehicle-collisions.csv', col_types = list(DATE = col_character())) # NYPD collision data 
population <- read_csv('data/New_York_City_Population_By_Neighborhood_Tabulation_Areas.csv') # US census of population by NTA 
```

```{r bikepath_read, echo=FALSE, message=FALSE, results='hide'}
# read in shapefile
bike_network <- read_sf('data/20180906_current_bike_network/20180906_current_bike_network.shp')

# convert to sf and transform to web map projection  
bike_network_trans <- st_transform(bike_network, 4326)

bike_network_trans <- bike_network_trans %>% 
  mutate(labs = paste0('<strong>Street: </strong>', street,
                       '<br><strong>Installed: </strong>', instdate,
    '<br><strong>Modified: </strong>', moddate,
    '<br><strong>Lanes: </strong>', lanecount,
    '<br><strong>Path Type Endpoint: </strong>', ft_facilit,
    '<br><strong>Path Type Startpoint: </strong>', tf_facilit))

labs <- as.list(bike_network_trans$labs)
```

New York City has seen an [**exponential increase**](http://www.nyc.gov/html/dot/downloads/pdf/cycling-in-the-city.pdf) in bicycle usage in the last 20 years. It reports [**up to 76% fewer**](https://www1.nyc.gov/html/dot/downloads/pdf/nyc-cycling-risk.pdf) injuries or deaths per 10 million miles biked since 2000, a measure of bike safety that controls for increased usage. Even so, the city has had an [**exceptionally deadly year**](https://www.nytimes.com/2019/08/21/nyregion/bike-deaths-nyc.html) for cyclists, leading many to feel the city needs to do more. There is [**existing evidence**](https://ajph.aphapublications.org/doi/full/10.2105/AJPH.2012.300762?journalCode=ajph) that (protected) biking infrastructure is an effective risk-reduction method. Some questions I seek to address:

* What is the breakdown of biking infrastructure in New York City?
* Where are injuries and deaths happening?
* Is there a relationship between bike infrastructure (or lack thereof) and cyclist injuries and deaths?
* Are there neighborhoods with less access to (protected) bike infrastructure? 
  + If so, can they be predicted by socioeconomic status?

## Let's get a quick understanding of bikes in New York.   

```{r date_manipulation, echo=FALSE}
# convert date to a date variable
NYPD$DATE <- NYPD$DATE %>% 
  substr(start = 1, stop = 10) %>% 
  ymd()

# create summarized subset of NYPD data 
sub <- NYPD 

# summarize by week
sub <- sub %>% 
  filter(DATE > 16438) %>% 
  mutate(week = floor_date(DATE, 'weeks', week_start = 7)) %>% 
  group_by(week) %>% 
  summarize(cyclist_injuries = sum(`NUMBER OF CYCLIST INJURED`))

```

### When do cyclist injuries and accidents happen?

<br/>

```{r}
heatmap <- NYPD %>% 
  mutate(day = wday(DATE)) %>% 
  mutate(hour = hour(TIME)) %>% 
  group_by(day, hour) %>% 
  summarize(injuries = sum(`NUMBER OF CYCLIST INJURED`)) %>% 
  ungroup() %>% 
  mutate(day_char = as.factor(day)) %>% 
  mutate(day_char = factor(day_char, ordered = TRUE, labels = c('Sunday', 'Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday', 'Saturday')))

ggplot(data = heatmap, aes(x = hour, y = day_char, fill = injuries)) + 
  geom_tile() + 
  scale_y_discrete(limits = rev(levels(heatmap$day_char))) + 
  scale_x_continuous(breaks = rep(0:23)) +
  labs(title = 'Most accidents occur during evening rush hour', fill = 'Injuries') + 
  scale_fill_viridis_c() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank()
        )
```


### What about historic trends?

<br/>

```{r injuries_plotly, echo=FALSE, message=FALSE, out.width='100%'}
# plot cyclist injuries per week 
plt.sub <- sub %>% 
  ggplot(aes(x = week, y = cyclist_injuries)) +
  geom_point(color = 'steelblue4', alpha = .7) + 
  geom_smooth(method='lm') + 
  xlab('Week') +
  ylab('Cyclist Injuries Per Week') + 
  theme_classic()

# convert to plotly object
ggplotly(plt.sub) %>% 
  layout(title = list(text = 'Cyclist injuries due to vehicle collisions are increasing.',
                      font = list(
                        size = 16
                        )
                      )
         ) %>% 
  config(displayModeBar = FALSE, scrollZoom = FALSE)
```

```{r}
# plot_ly(sub,
#         x = ~week,
#         y = ~cyclist_injuries,
#         marker = list(color = 'steelblue4', alpha = .7)
#         ) %>% 
#   layout(title = list(text = 'Cyclist injuries due to vehicle collisions are increasing.',
#                       font = list(size = 16
#                                   )
#                       ), 
#          xaxis = 'Year',
#          yaxis = 'Cyclist Injuries Per Week'
#          ) %>% 
#   config(displayModeBar = FALSE, scrollZoom = FALSE)
```

### What kinds of bike paths are available in the city?

<br/>

```{r path_types}
path_type <- bike_network_trans 

# convert NA path types to explicit factor level
path_type$ft_facilit <- fct_explicit_na(path_type$ft_facilit)

# Collapse small and NA path types into Other 
path_type$ft_facilit <- fct_collapse(path_type$ft_facilit, 
                                     Other = c('Link',
                                               'Sidewalk',
                                               'Ped Plaza',
                                               'Dirt Trail',
                                               'Buffered Conventional',
                                               '(Missing)'
                                               )
                                     )


path_type <- path_type %>% 
  mutate(year_installed = year(instdate)) %>%
  group_by(ft_facilit, year_installed) %>% 
  arrange(ft_facilit, year_installed) %>% 
  ungroup() %>%
  group_by(ft_facilit) %>%
  # Get cumulative sum of miles by path type
  mutate(cum_miles = cumsum(LaneMiles)) %>% 
  ungroup() %>% 
  group_by(ft_facilit, year_installed) %>% 
  # Summarize into path type x year summary with cumulative sum of miles
  summarize(count = n(), miles = round(max(cum_miles), 2)) %>%
  filter(year_installed > 2000) %>%
  as_tibble() %>% 
  select(-geometry)

# Manually re-define factors to put Other at bottom
path_type$ft_facilit <- path_type$ft_facilit %>% 
  factor(levels = c('Bike-Friendly Parking', 
                    'Curbside',
                    'Greenway',
                    'Protected Path',
                    'Sharrows',
                    'Signed Route',
                    'Standard',
                    'Other'
                    )
         )
# Dataset is missing Greenway installations in 2018 (or there may not be any. Manually add 2017 max)
man_add <- tibble(ft_facilit = factor('Greenway',
                                      levels = c('Bike-Friendly Parking',
                                                 'Curbside',
                                                 'Greenway',
                                                 'Protected Path',
                                                 'Sharrows',
                                                 'Signed Route',
                                                 'Standard',
                                                 'Other'
                                                 )
                                      ),
                  year_installed = 2018,
                  count = 0,
                  miles = 309.13
                  )

path_type <- path_type %>% 
  dplyr::bind_rows(man_add)
  

path_type <- path_type %>% 
  arrange(ft_facilit, year_installed)

# Stacked bar chart of miles of path type by year
path_type_plot <- ggplot(path_type) +
  geom_bar(aes(x = year_installed, y = miles, fill = ft_facilit), position = 'stack', stat = 'identity') + 
  theme_classic() +
  labs(x = 'Year', y = 'Miles', fill = 'Path Type') + 
  theme(legend.title = element_blank()) + 
  scale_fill_viridis_d()

# convert to plotly object
ggplotly(path_type_plot) %>% 
  layout(title = list(text = 'Bike paths have increased exponentially in the last 20 years.',
                      font = list(
                        size = 16
                        )
                      )
         ) %>% 
  config(displayModeBar = FALSE, scrollZoom = FALSE)
```


### Where do cyclist injuries and deaths occur?   

<br/>

```{r leaflet, echo=FALSE, warning=FALSE, layout='l-screen-inset', fig.height=5}
# preferCanvas to improve performance with high volume of markers
map <- leaflet(data = NYPD, options = leafletOptions(minZoom = 11, maxZoom = 18, preferCanvas = TRUE)) %>%

  enableTileCaching() %>%
  
  # updateWhenZooming and updateWhenIdle to improve performance 
  addTiles(group = 'Color', options = tileOptions(updateWhenZooming = FALSE, updateWhenIdle = TRUE)) %>% 
  addProviderTiles(providers$Stamen.TonerLite, group = 'Grayscale') %>% 


  # restrict boundaries to around NYC (doesn't seem to work as expected)
  fitBounds(lng1 = min(NYPD$LONGITUDE) - 0.11, 
            lat1 = min(NYPD$LATITUDE) - 0.11,
            lng2 = max(NYPD$LONGITUDE) + 0.11, 
            lat2 = max(NYPD$LATITUDE) + 0.11) %>% 
  
  # set default view to NYC
  setView(lng = -73.94, lat = 40.729, zoom = 11) %>%  
  
  # add markers where a cyclist was injured or killed, define popup html 
  addMarkers(clusterOptions = markerClusterOptions(),
             data = NYPD[NYPD$`NUMBER OF CYCLIST INJURED` > 0 | NYPD$`NUMBER OF CYCLIST KILLED` > 0,],
             ~LONGITUDE,
             ~LATITUDE,
            popup = ~ paste0('<strong>','Date: ', '</strong>', DATE, '<br/>',
                             '<strong>', 'Time: ', '</strong>', TIME, '<br/>',
                             '<strong>', 'Injured: ', '</strong>', `NUMBER OF CYCLIST INJURED`, '<br/>',
                             '<strong>', 'Killed: ', '</strong>', `NUMBER OF CYCLIST KILLED`, '<br/>',
                             '<strong>', 'Vehicle Type 1: ', '</strong>', `VEHICLE TYPE CODE 1`, '<br/>',
                             '<strong>', 'Cause - Vehicle 1: ', '</strong>', `CONTRIBUTING FACTOR VEHICLE 1`, '<br/>', 
                             '<strong>', 'Vehicle Type 2: ', '</strong>', `VEHICLE TYPE CODE 2`, '<br/>',
                             '<strong>', 'Cause - Vehicle 2: ', '</strong>', `CONTRIBUTING FACTOR VEHICLE 2`)) %>% 
  
  # add bike path network, define labels and tooltip options for continuous paths                           
  addPolylines(data = bike_network_trans,
              color = 'darkblue',
              weight = 3,
              opacity = 1,
              smoothFactor = .5,
              label = lapply(labs, HTML),
              highlightOptions = highlightOptions(color = 'red',
                                                  weight = 4, 
                                                  bringToFront = TRUE)) %>% 
  addResetMapButton() %>% 
  
  # Add user controls to toggle groups displayed
  addLayersControl(
  baseGroups = c('Color', 'Grayscale'),
  options = layersControlOptions(collapsed = TRUE)
  ) %>%
  
  addControl("<P style='font-size:10px'>
              <B>Hint!</B>
              <br> Zoom in and click on individual points for accident details.
              <br> Hover over blue lines to see bike path details.
              </P>",
  position='bottomleft')

map
```

### An alternative map implemented in Deck.gl: 

<br/>

```{r mapdeck, layout='l-page', fig.height=4, out.width='100%', fig.align='center'} 
set_token(Sys.getenv('MAPBOX')) # must use your own Mapbox Access Token if reproducing

mapdeck(style ='mapbox://styles/jbixon/cjxchrj87086s1ct8cs8m351s', # customized map with buildings added
        location = c(-73.987341, 40.724758),
        bearing = 14.9,
        zoom = 12.1,
        pitch = 45) %>% 
  add_path(data = bike_network_trans,
           layer_id = 'paths',
           auto_highlight = TRUE,
           stroke_colour = 'ft_facilit',
           tooltip = 'labs',
           stroke_width = 3
           ) %>% 
  add_scatterplot(data = NYPD[NYPD$`NUMBER OF CYCLIST INJURED` > 0 | NYPD$`NUMBER OF CYCLIST KILLED` > 0,],
                  layer_id = 'points',
                  lon = 'LONGITUDE',
                  lat = 'LATITUDE',
                  radius = 20,
                  radius_min_pixels = 2,
                  radius_max_pixels = 25,
                  fill_opacity = .7,
                  )
```

```{r lookup_table, echo=FALSE}
 
year <- rep(2012:2018)
nta <- rep(1:195)

# create lookup table with patterns for key creation (every NTA repeated for every year)
lookup <- data.frame(nta = rep(1:195, times = 7), year = rep(year, each = 195)) %>% 
  mutate(key = paste0(nta, '-', year)) %>% 
  dplyr::arrange(nta, year)
```

```{r population_processing, echo=FALSE}
# prep population dataset for joining to nyc_poly
population <- population %>% 
  filter(Year == 2010) %>% 
  select(`NTA Code`, Population) %>% 
  rename(nta_id = `NTA Code`)
```

```{r point_aggregation_and_joining, echo=FALSE}
# get Neighborhood Tabulation Area boundaries and join to population by NTA
nyc_poly <- nyc_boundaries(geography = "nta") %>%
  st_transform(4326) %>% 
  mutate(nta = rep(1:195)) %>% 
  left_join(population, by = 'nta_id')

# prep accident data for spatial operations (keep only geotagged & cyclist accidents)
NYPD_sum <- NYPD %>%
  filter(LATITUDE != is.na(TRUE)) %>%
  filter(`NUMBER OF CYCLIST INJURED` > 0) %>%
  st_as_sf(coords = c('LONGITUDE', 'LATITUDE'), crs = 4326)

# identify which accidents (points) intersect with NTAs (polygons)
# think it's giving me the indexes of accidents for each nta, need to figure out how to get date info from index
#lengths(st_intersects(nyc_poly, tst))

# try reversing the process: instead of count of intersections for each polygon, do st_intersects for each point so that each point is assigned a polygon it intersects with, then group_by day and polygon
#head(st_intersects(tst, nyc_poly))

NYPD_sum <- NYPD_sum %>%
  mutate(nta = as.numeric(st_intersects(NYPD_sum, nyc_poly))) %>%
  # floor accidents by year and clean dates for further joins and processing
  mutate(year = substr(floor_date(DATE, unit = 'year'), start = 1, stop = 4)) %>%
  group_by(nta, year) %>%
  # get count of accidents per year and NTA 
  summarize(accident_count = n()) %>% 
  mutate(key = paste0(nta, '-', year))

# join accident counts to lookup table (to solve problem of rows with 0 accidents being dropped)
join <- lookup %>% 
  left_join(NYPD_sum, by = 'key') %>% 
  arrange(nta.x, year.x) %>% 
  mutate(accident_count = if_else(is.na(accident_count), 0, as.double(accident_count))) %>% 
  select(-nta.y, -year.y) %>% 
  rename(nta = nta.x) %>% 
  rename(year = year.x)

# have summary by nta and month, but geometry is list of points, not polygon of nta. Need to replace with polygon for map visualization
join_poly <- join %>% 
  inner_join(as.data.frame(nyc_poly), by = 'nta') %>% 
  as.data.frame() %>% 
  select(-geometry.x) %>% 
  rename(geometry = geometry.y) %>% 
  # calculate accidents per capita from accidents and population per NTA 
  mutate(accidents_per_capita = accident_count / Population) %>% 
  # orders of magnitude difference, try square root transform
  mutate(accidents_per_capita_sqrt = sqrt(accidents_per_capita)) %>%
  # remove unnecesary variables, reorder remaining vars
  select(-state_fips, -county_fips, -county_name, -borough_id, -puma_id, -puma_name) %>% 
  select(nta, year, key, nta_id, nta_name, borough_name, accident_count,
         Population, accidents_per_capita, accidents_per_capita_sqrt, geometry) %>% 
  st_as_sf()
```

<br/> 

My first attempt to understand accidents spatially led me to aggregate them within the Neighborhood Tabulation Area (NTA) they occur in and divide by NTA population. This shows not just which neighborhoods have the most accidents, but which have the most **per capita**. It shows more clearly where accidents are happening than by mapping each individual accident, but it does not say whether a neighborhood is over-indexed on accidents from an infrastructure perspective. Neighborhood population may not be insightful because it may have little to do with how many people bike there, or the kind of infrastructure they have, or traffic speeds. I will explore this further in the future. 

```{r animated_map, echo=FALSE, layout='l-screen-inset'}
# create chloropleth map with accidents per capita fill, change graphics options for 1080p export
map_base <- ggplot(join_poly) +
  geom_sf(aes(fill = accidents_per_capita_sqrt)) +
  scale_fill_viridis_c() +
  theme_void() +
  theme(plot.title = element_text(size = 30)) +
  theme(plot.subtitle = element_text(size = 22)) +
  theme(legend.key.size = unit(1, 'in')) + 
  theme(legend.title = element_text(size = 24)) + 
  theme(legend.text = element_text(size = 20)) +
  theme(panel.grid = element_line(color = "transparent")) 

# animate by year, identify current year in title
anim <- map_base +
  labs(title = 'Cyclist injuries per capita per year in NYC', subtitle = 'Year: {closest_state}') +
  transition_states(year)

anim <- animate(anim, width = 1920, height = 1080)

anim
```

I then wanted to understand one of my primary questions: whether bike infrastructure could predict cyclist accidents. To do so, I combined the [**NYPD Motor Vehicle Collisions**](https://data.cityofnewyork.us/Public-Safety/NYPD-Motor-Vehicle-Collisions-Crashes/h9gi-nx95) and [**NYC Street Centerline**](https://data.cityofnewyork.us/City-Government/NYC-Street-Centerline-CSCL-/exjm-f27b) datasets to get a distinct count of bike accidents for every street segment as a dependent variable. 

Two [**simple and proven**](https://www.researchgate.net/publication/222659783_The_Statistical_Analysis_of_Crash-Frequency_Data_A_Review_and_Assessment_of_Methodological_Alternatives) models for regressing discrete count data are the poisson regression and negative-binomial regression. They are not without their flaws, but they are a good starting point.

```{r accident_distance_calc}
zip <- unzip('data/NYC_Street_Centerline_CSCL.zip', 
             exdir = '/c/Users/jbjrV/OneDrive/Radix-blog/_posts/2019-01-24-mapping-cyclist-safety-in-new-york-city/data')

# use NAD83 crs for later distance operation
streets <- read_sf(zip) %>% 
  st_transform(3488)

accidents <- NYPD %>% 
  filter(LATITUDE != is.na(TRUE)) %>% 
  filter(`NUMBER OF CYCLIST INJURED` > 0 | `NUMBER OF CYCLIST KILLED` > 0) %>% 
  st_as_sf(coords = c('LONGITUDE', 'LATITUDE')) %>% 
  st_set_crs(4326)

# increase street width to assign accidents that fall within  
streets_buffer <- streets %>% 
  st_buffer(dist = 10) # 10 meter buffer seems to be the right size for NYC

# transform to web mercator to align crs with accidents for intersect operation
streets_buffer <- streets_buffer %>% 
  st_transform(4326)  

intersect <- st_intersects(streets_buffer, accidents)

# get count of accidents that intersect with each street - one row per street
intersect_count <- st_intersects(streets_buffer, accidents) %>% 
  lengths() %>% 
  enframe()

# append accident count onto original street table
streets_accidents <- streets %>% 
  bind_cols(intersect_count) %>% 
  select(-name) %>% 
  rename(accidents = value) %>%
  mutate(bike_lane = fct_explicit_na(bike_lane)) %>%  
  # mutate(bike_lane = factor(bike_lane,
  #                           labels = c('Class I',
  #                                      'Class II',
  #                                      'Class III',
  #                                      'Links',
  #                                      'Class I, II',
  #                                      'Class II, III',
  #                                      'Stairs',
  #                                      'Class I, III',
  #                                      'Class II, I',
  #                                      'Class III, I',
  #                                      'Class III, II'
  #                                      )
  #                           )
  #        ) %>% 
  st_as_sf()
```

```{r model}
# poisson vs. negative binomial regression
# dispersion is when variance > mean - NB is better for this 
# when dispersion statistic is close to one, use poisson
# when > 1, use negative binomial 

p <- glm(accidents ~ bike_lane, family = 'poisson', data = streets_accidents)

nb <- MASS::glm.nb(accidents ~ bike_lane, data = streets_accidents)

sjPlot::tab_model(p,
                  nb,
                  dv.labels = c('Poisson', 'Negative-Binomial'),
                  title = 'Predicting Accident Count',
                  p.style = 'asterisk'
                  )
```


```{r notes}
#introduction - need to redo
# trend scatter plot - done, maybe look into time-series analysis
# gif of NTAs - have accident counts, change to accidents/NTA population (and acknowledge issues with using 2010 survey data but it's better than a simple population map)
# bar plot of accident causes - not done
# leaflet - done, consider any simplifications but likely done
# model predicting bike path/path type - not done 
# aadt traffic / volume measurements of traffic 
# look at how cities measure volume or traffic for cars and emulate for bikes
# strava geolocation for bike data ??
# conclusions - not done
# seperate post about point aggregation in chloropleths - not until all else done 
```

***

### Note:    

This is an ongoing project that continues to evolve as I learn. If you have any constructive criticism please feel free to reach out, especially with suggestions about traffic accident modeling methodology.  

    

